---
title: "KingClass1Lab"
author: "Bernadette King"
date: "2025-08-28"
output: html_document
---
```{r}
library(dplyr)
library(haven)
library(readr)
library(openxlsx)
library(tidyverse)
```
#1. Import class one survey data from our Github site (https://github.com/kijohnson/Advanced-Data-Analysis) (The dataset is called ‘Class 1 Survey Fall 2025 2.csv’), calling the R dataframe that you create C1survey.
```{r}
#inputting the data
C1survey <- read_csv ("https://github.com/kijohnson/Advanced-Data-Analysis/raw/refs/heads/main/Class%201%20Survey%20Fall%202025%202.csv")

```

#2. Determine: a. the number of observations (i.e. the number of people who filled out the survey) and b. the number of variables in the dataframe.
```{r}
#displaying the number of variables and observations
dim(C1survey)
```

#3a. Rename the column variables to something shorter and that is descriptive of what the variable is about (for example like_dogs for the ‘Do you like dogs?’ question variable) and 
#b. Display that they are renamed. Hints: You can use either the rename_with function or the names function. The list of names for time is:

“id”, “like_cats”, “like_dogs”, “have_desert”, “slogan”, “fav_day”, “larkORowl”, “fav_food”, “fav_drink”, “fav_season”, “fav_month”, “hobby”, “program”, “specialization”, “stat_software”, “R_exp”, “coding_comfort”, “coding_length”, “top_three”, “public_health_interest”, “fav_num”, “bday”, “bmonth”, “country”, “state”, “city”, “highest_educ_level”

```{r}
#renaming the survey question variables
names(C1survey)[1:27] <- c("id", "like_cats", "like_dogs", "have_desert", "slogan", "fav_day", "larkORowl", "fav_food", "fav_drink", "fav_season", "fav_month", "hobby", "program", "specialization", "stat_software", "R_exp", "coding_comfort", "coding_length", "top_three", "public_health_interest", "fav_num", "bday", "bmonth", "country", "state", "city", "highest_educ_level")

#displaying the new names 
View(C1survey)

```

#4. Write code to determine and display the number of factor, integer, numerical, and character variables there are in the C1survey dataset. Suggestion: Try to ask ChatGPT for a solution if you are unsure. There is a really short answer to this but we want to see what students come up with!

```{r}
#creating a table to display number of each class of variables
table(sapply(C1survey, class))

```

#5a. Using code, check the bday and bmonth variables for any unusual or missing values (hint: use table function). If you find any, b. describe what you will do with the unusual values in a comment before or immediately following the code that does it, and c. after you have cleaned up any unusual values, find the median bday and month. *Note: please make sure that your bday and month variables are numeric before doing statistics

```{r}
#viewing the data for birthdays
table(C1survey$bday)

#converting incorrect bday format to numeric format
C1survey <- C1survey %>%
  mutate(
    bday= as.numeric(str_extract(bday, "\\d+"))
  )
#viewing the data for birthday month
table(C1survey$bmonth)

#converting character months to numeric months
C1survey <- C1survey %>% 
  mutate(bmonth = case_when(bmonth == "August" ~ 1,
                            bmonth == "December" ~ 12,
                            bmonth == "January" ~ 1,
                            bmonth == "june" ~ 6,
                            bmonth == "June" ~ 6, 
                            bmonth == "May" ~ 5,
                            bmonth == "November" ~ 11,
                            bmonth == "October" ~ 10,
                            bmonth == "Septemeber" ~ 9,
                            TRUE ~ as.numeric(bmonth))
)
C1survey
```
#6a. Create a new variable called bseason that gives the season according to Northern Meteorological season in which respondents were born (winter = 12/1 to 2/29, spring = 3/1 to 5/31, summer = 6/1 to 8/31, fall = 9/1 to 11/30). b. Using the table function, print a table of Seasons in the columns and bmonths in the rows that allows you to check that the coding is correct c. Sum the columns using the addmargins function to answer the question of how many classmates were born in each season? Include your answer in the code.
```{r}
# Create bseason based on bmonths using Northern Meteorological seasons
bseason <- ifelse(C1survey$bmonth %in% c(12, 1, 2), "Winter",
           ifelse(C1survey$bmonth %in% c(3, 4, 5), "Spring",
           ifelse(C1survey$bmonth %in% c(6, 7, 8), "Summer",
           "Fall")))

# Create a cross-tabulation of months (rows) and seasons (columns)
season_table <- table(C1survey$bmonth, bseason)
print(season_table)

# Add margins to get total number of births per season
season_totals <- addmargins(season_table, margin = 2)  # margin = 2 => add column totals
print(season_totals)

```
#7. Pick your favorite variable or variables to analyze, come up with a question you want to answer with that variable or variables, generate the code, and provide an answer to your question. Describe what you did using comments (i.e. #’s).

```{r}
# Looking at the larkORowl variable, how many individuals selected owl, lark or hummingbird?

#simplfying the variable names 
C1survey <- C1survey %>% 
  mutate(larkORowl = case_when(larkORowl == "Lark ( I am a morning person)" ~ "Lark",
                            larkORowl == "Hummingbird (I am neither a lark nor an owl)" ~ "Hummingbird",
                            larkORowl == "Owl (I am a night person)" ~ "Owl"))

#Looking at the amount of each response (why won't it work without evaluating the variable first?):

# Check structure
str(C1survey$larkORowl)

# Check the unique values
unique(C1survey$larkORowl)

# Check the first few values
head(C1survey$larkORowl)

# Then try to count again
table(C1survey$larkORowl)




```







